---
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
params:
  DataDate:    "2024-05-25"
  FirstYear:   2014
  LastYear:    2025
  TargetArea:  "Kansas"
title: "Help America Vote Verification (HAVV) Data -- `r params$TargetArea`"
author: "Earl F Glynn<br><small>watchdoglab.substack.com</small>"
date: "<small>`r Sys.Date()`</small>"
---

```{r TableOfContents, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Backstage {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r}
EXCEL_LIMIT <- 2^20
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)   # date functions
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

Graphics

```{r Graphics}
library(scales)      # comma, comma_format
library(ggpubr)
```

Utilities

```{r Utilities}
library(downloader)  # download
library(tools)       # md5sum
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(
                  axis.text             = element_text(size = 12),
                  axis.title            = element_text(size = 14),

                  legend.position       = "bottom",

                  plot.caption          = element_text(hjust = c(0.0, 1.0),
                                                       size = 8),
                  plot.caption.position = "plot",

                  plot.title.position   = "plot",
                  plot.title            = element_text(size = 14),
                  plot.subtitle         = element_text(size = 12, color = "blue"),

                  strip.background      = element_rect(fill = "aliceblue"),
                  strip.text            = element_text(size = 14),

                  title                 = element_text(size = 14))
               )

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width)
}
```

# Setup

```{r Caption}
CAPTION <- paste("Source:  Social Security Help America Vote Verification (HAVV) Transactions by State\nEntire Dataset,",
                 params$DataDate)

stamp   <- paste("efg", format(Sys.time(), "%Y-%m-%d"))
```

# Load weekly data

```{r LoadData}
havvWeekly <-
  read_xlsx(paste0(params$DataDate, "/HAVV-Through-", params$DataDate, ".xlsx"))  |>
  filter(
          !is.na(Year),                 # Exclude yearly total lines
          State_Territory != "Total"    # Exclude weekly total lines
        )
dim(havvWeekly)
```

```{r StateCounts}
stateCounts <- table(havvWeekly$State_Territory)
length(stateCounts)
```
TargetArea

```{r TargetArea}
targetArea <-
  havvWeekly                            |>
  filter(State_Territory == params$TargetArea,
         !is.na(Date),
         Year >= params$FirstYear,
         Year <= params$LastYear)       |>
  arrange(Sheet)                       
dim(targetArea)
```

# Total Transactions

```{r SetRange}
startDate <- as.POSIXct(paste0(params$FirstYear, "-01-01"))
endDate   <- startDate + years(params$LastYear - params$FirstYear)
```

```{r TotalTransactions, fig.width = 8, fig.height = 3.5, warning = FALSE}
plotTransactions <- 
  ggplot(targetArea, aes(x = Date, y = Total_Transactions))     +
  geom_step(size = 1.5, color = "blue")                         +

  scale_x_datetime(
                    date_labels  = "%Y",                        # "%Y\n%m"
                    date_breaks  = "1 year",
               #    date_minor_breaks = "3 months",             # These DO NOT align correctly. Why?
                    limits       = c(startDate, endDate)
               #    expand = expansion(0, 0)                    # This solution for breaks, truncates axis labels
                  )                                                          +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::comma)                                 +
  geom_vline(xintercept = seq(from = startDate, to = endDate, by = "year" ),
               color = "grey", linetype = "dashed", size = 1)                +
  labs(
         title = paste0("HAVV Data for ", params$TargetArea, ":  Total Transactions"),
         x = "Week",
         y = "Total Transactions",
         caption = c(CAPTION, stamp)
       )                                                                     +
  theme(
          panel.grid.minor.x = element_blank(),  # these do not align correctly
          panel.grid.minor.y = element_blank(),  # these are not wanted

       )

print(plotTransactions)
```

# Percent NonMatches

```{r PercentNonMatch, fig.width = 8, fig.height = 3.5, warning = FALSE}
plotNonMatch <-
  ggplot(targetArea, aes(x = Date, y = PercentNonMatch))        +
  geom_step(size = 1.5, color = "red")                          +

  scale_x_datetime(
                    date_labels  = "%Y",                        # "%Y\n%m"
                    date_breaks  = "1 year",
               #    date_minor_breaks = "3 months",             # These DO NOT align correctly. Why?
                    limits       = c(startDate, endDate)
               #    expand = expansion(0, 0)                    # This solution for breaks, truncates axis labels
                  )                                                          +
  scale_y_continuous(labels = scales::comma)                                 +
  geom_vline(xintercept = seq(from = startDate, to = endDate, by = "year" ),
               color = "grey", linetype = "dashed", size = 1) +
  labs(
         title = paste0("HAVV Data for ", params$TargetArea, ":  Percent Nonmatching Transactions"),
         x = "Week",
         y = "Percent Nonmatching",
         caption = c(CAPTION, stamp)
       )                                                                     +
  theme(
          panel.grid.minor.x = element_blank(),  # these do not align correctly
          panel.grid.minor.y = element_blank(),  # these are not wanted

       )

print(plotNonMatch)
```

# Percent Deceased in Matching Transactions

```{r PercentMatchDeceased, fig.width = 8, fig.height = 3.5, warning = FALSE}
plotMatchDeceased <-
  ggplot(targetArea, aes(x = Date, y = PercentMatchDeceased))     +
  geom_step(size = 1.5, color = "black")                          +

  scale_x_datetime(
                    date_labels  = "%Y",                        # "%Y\n%m"
                    date_breaks  = "1 year",
               #    date_minor_breaks = "3 months",             # These DO NOT align correctly. Why?
                    limits       = c(startDate, endDate)
               #    expand = expansion(0, 0)                    # This solution for breaks, truncates axis labels
                  )                                                          +
  scale_y_continuous(labels = scales::comma )                                +
  geom_vline(xintercept = seq(from = startDate, to = endDate, by = "year" ),
               color = "grey", linetype = "dashed", size = 1.25) +
  labs(
         title = paste0("HAVV Data for ", params$TargetArea, ":  Percent Deceased in Matching Transactions"),
         x = "Week",
         y = "Matching Percent Deceased",
         caption = c(CAPTION, stamp)
       )                                                                     +
  theme(
          panel.grid.minor.x = element_blank(),  # these do not align correctly
          panel.grid.minor.y = element_blank(),  # these are not wanted

       )

print(plotMatchDeceased)
```

# Three grapics in one

```{r CreatePDF, fig.width = 8.5, fig.height = 11}
pdf(paste0("HAVV-Data-", params$DataDate, ".pdf"),
    width = 8.5, height = 11)
 
figure <- 
  ggarrange(plotTransactions, plotNonMatch, plotMatchDeceased,
            labels = c("", "", ""),
            ncol = 1, nrow = 3)
print(figure)

dev.off()
```

```{r CreatePNG, fig.width = 8.5, fig.height = 11}
png(paste0("HAVV-Data-", params$DataDate, ".png"),
    units = "in", width = 8.5, height = 11, pointsize = 12, res =300)
 
figure <- 
  ggarrange(plotTransactions, plotNonMatch, plotMatchDeceased,
            labels = c("", "", ""),
            ncol = 1, nrow = 3)
print(figure)

dev.off()
```

# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

